Inferring the rate of Earth-like exoplanets from noisy data

### 2014-03-23 document/ms.tex
starting document

% has to be before amssymb it seems
\usepackage{color,hyperref}
\definecolor{linkcolor}{rgb}{0,0,0.5}
\hypersetup{colorlinks=true,linkcolor=linkcolor,citecolor=linkcolor,
            filecolor=linkcolor,urlcolor=linkcolor}

\usepackage{url}
\usepackage{algorithmic,algorithm}
\usepackage{multirow}

\usepackage{listings}
\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
\lstset{language=Python,
        basicstyle=\footnotesize\ttfamily,
        showspaces=false,
        showstringspaces=false,
        tabsize=2,
        breaklines=false,
        breakatwhitespace=true,
        identifierstyle=\ttfamily,
        keywordstyle=\bfseries\color[rgb]{0.133,0.545,0.133},
        commentstyle=\color[rgb]{0.133,0.545,0.133},
        stringstyle=\color[rgb]{0.627,0.126,0.941},
    }

\input{vars}

@@

\title{%
Inferring the exoplanet radius distribution from a noisy, incomplete catalog
}

\newcommand{\nyu}{2}
\newcommand{\mpia}{3}
\author{%
    Daniel~Foreman-Mackey\altaffilmark{1,\nyu},
    David~W.~Hogg\altaffilmark{\nyu,\mpia},
    \etal
}
\altaffiltext{1}{To whom correspondence should be addressed:
                        \url{danfm@nyu.edu}}
\altaffiltext{\nyu}{Center for Cosmology and Particle Physics,
                        Department of Physics, New York University,
                        4 Washington Place, New York, NY, 10003, USA}
\altaffiltext{\mpia}{Max-Planck-Institut f\"ur Astronomie,
                        K\"onigstuhl 17, D-69117 Heidelberg, Germany}

\begin{abstract}
The catalog of \kepler\ Objects of Interest (KOI) is a list of putative periodic exoplanet transit signals
  discovered in the photometric data from the \kepler\ Satellite.
In principle this catalog contains a wealth of statistical information about the full population of exoplanets.
Here we construct a model of the population that explains the content of the catalog.
Our approach is novel in that
  it takes the form of a rigorous (if approximate) hierarchical Bayesian inference,
  it includes a parameterization of a non-trivial selection function (incompleteness),
  and we fully marginalize out all latent and nuisance parameters using sampling.
Because the model is properly hierarchical,
  it can be constrained by multiple data sources;
  we demonstrate this by adding the Solar System as a constraint,
  but extensions to radial velocity, direct detection, and astrometric surveys would also be straightforward.
Approximations abound:
  Simple parameterized forms are assumed for the multivariate distribution of exoplanet parameters,
  incompleteness is assumed to depend only on a scalar related to signal-to-noise,
  and the KOI catalog is assumed to be dominated by real (not false-positive) exoplanet signals.
We find XXX and YYY.
\end{abstract}

\keywords{% CUT THESE DOWN TO SIX
  astronomical~databases:~miscellaneous
  ---
  catalogs
  ---
  eclipses
  ---
  methods:~statistical
  ---
  planetary~systems
  ---
  planets~and~satellites:~detection
  ---
  planets~and~satellites:~fundamental~parameters
  ---
  stars:~statistics
  ---
  surveys
}

@@

\begin{figure}[htbp]
\begin{center}
    \includegraphics{gm.pdf}
\end{center}
\caption{%
\figlabel{gm}}
\end{figure}

% \begin{deluxetable}{ccl}
% \tablecaption{%
% A description of the model parameters.
% \tablabel{parameters}}
% \tablewidth{0pt}
% \tablehead{& & \colhead{Description}}
% \startdata
% \multirow{3}{*}{\population}
% & $\population_\period$ & The period distribution \\
% & $\population_\radius$ & The exoplanet radius distribution \\
% & $\population_\relincl$ & The relative inclination distribution \\
% \hline
% \selection & --- & The parameters of the catalog selection function \\
% \hline
% $\isobs_{kn}$ & --- & A flag indicating the presence of exoplanet $k$ orbiting
% star $n$ in the catalog \\
% \hline
% \multirow{3}{*}{$\planet_{kn}$}
% & $\period_{kn}$ & The period of exoplanet $k$ orbiting star $n$ (or null) \\
% & $\radius_{kn}$ & The exoplanet's radius \\
% & $\relincl_{kn}$ & The inclination of this orbit away from the systemic mean \\
% \hline
% \multirow{3}{*}{$\planetobs_{kn}$}
% & $\periodobs_{kn}$ & The observed exoplanet period and uncertainty \\
% & $(\rorobs)_{kn}$ & The observed radius ratio and uncertainty \\
% & $\impactobs_{kn}$ & A constraint on the observed impact parameter \\
% \hline
% \multirow{4}{*}{$\stellar_{n}$}
% & $\smass_{n}$ & The mass of star $n$ \\
% & $\sradius_{n}$ & The radius of star $n$ \\
% & $\snoise_{n}$ & The ``variability'' of star $n$ \\
% & $\incl_{n}$ & The mean inclination of the exoplanets in the system $n$ \\
% \hline
% \multirow{2}{*}{$\stellarobs_{n}$}
% & $\sloggobs_{n}$ & The observed surface gravity (and uncertainty) of star $n$ \\
% & $\snoiseobs_{n}$ & The estimated variability of star $n$ \\
% \enddata
% \end{deluxetable}

\acknowledgments
SAMSI. %
It is a pleasure to thank
    \ldots
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).

@@

\bibitem[Fang \& Margot(2012)]{fang}
Fang, J., \& Margot, J.-L.\ 2012, \apj, 761, 92
\arxiv{1207.5250}

\bibitem[Tremaine \& Dong(2012)]{tremaine}
Tremaine, S., \& Dong, S.\ 2012, \aj, 143, 94
\arxiv{1106.5403}

>>>>

\usepackage{color,hyperref}
\definecolor{linkcolor}{rgb}{0,0,0.5}
\hypersetup{colorlinks=true,linkcolor=linkcolor,citecolor=linkcolor,
            filecolor=linkcolor,urlcolor=linkcolor}
\usepackage{url}
\usepackage{amssymb,amsmath}

\newcommand{\project}[1]{{\sffamily #1}}
\newcommand{\emcee}{\project{emcee}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\license}{MIT License}

\newcommand{\paper}{\emph{Article}}

\newcommand{\foreign}[1]{\emph{#1}}
\newcommand{\etal}{\foreign{et\,al.}}
\newcommand{\etc}{\foreign{etc.}}

\newcommand{\Fig}[1]{Figure~\ref{fig:#1}}
\newcommand{\fig}[1]{\Fig{#1}}
\newcommand{\figlabel}[1]{\label{fig:#1}}
\newcommand{\Tab}[1]{Table~\ref{tab:#1}}
\newcommand{\tab}[1]{\Tab{#1}}
\newcommand{\tablabel}[1]{\label{tab:#1}}
\newcommand{\Eq}[1]{Equation~(\ref{eq:#1})}
\newcommand{\eq}[1]{\Eq{#1}}
\newcommand{\eqlabel}[1]{\label{eq:#1}}
\newcommand{\Sect}[1]{Section~\ref{sect:#1}}
\newcommand{\sect}[1]{\Sect{#1}}
\newcommand{\App}[1]{Appendix~\ref{sect:#1}}
\newcommand{\app}[1]{\App{#1}}
\newcommand{\sectlabel}[1]{\label{sect:#1}}

\newcommand{\dd}{\ensuremath{\,\mathrm{d}}}
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}

% TO DOS
\newcommand{\todo}[3]{{\color{#2} \emph{#1} TODO: #3}}
\newcommand{\dfmtodo}[1]{\todo{DFM}{red}{#1}}
\newcommand{\hoggtodo}[1]{\todo{HOGG}{blue}{#1}}
\newcommand{\mortontodo}[1]{\todo{MORTON}{green}{#1}}

% Document specific variables.
\newcommand{\rate}{\ensuremath{\eta}}
\newcommand{\ratepars}{\ensuremath{\theta}}
\newcommand{\obs}[1]{\ensuremath{\hat{#1}}}
\newcommand{\radius}{\ensuremath{r}}
\newcommand{\period}{\ensuremath{T}}
\newcommand{\completeness}{\ensuremath{P_\mathrm{c}}}
\newcommand{\transitprob}{\ensuremath{P_\mathrm{t}}}

\newcommand{\interim}{\ensuremath{\alpha}}

@@

\title{%
    Inferring the rate of Earth-like exoplanets from noisy data
}

\newcommand{\nyu}{2}
\newcommand{\mpia}{3}
\newcommand{\princeton}{4}
\author{%
    Daniel~Foreman-Mackey\altaffilmark{1,\nyu},
    David~W.~Hogg\altaffilmark{\nyu,\mpia},
    Timothy~D.~Morton\altaffilmark{\princeton},
    \etal
}
\altaffiltext{1}            {To whom correspondence should be addressed:
                             \url{danfm@nyu.edu}}
\altaffiltext{\nyu}         {Center for Cosmology and Particle Physics,
                             Department of Physics, New York University,
                             4 Washington Place, New York, NY, 10003, USA}
\altaffiltext{\mpia}        {Max-Planck-Institut f\"ur Astronomie,
                             K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{\princeton}   {Department of Astrophysics, Princeton University,
                             Princeton, NJ}

\begin{abstract}

We infer the distribution and rate of Earth-like exoplanets around Sun-like
stars.

\end{abstract}

\keywords{%
exoplanets: sickness
---
code: open-source
---
keywords: made-up-by-Hogg
}

@@

\section{The model}

\emph{Note: I think that we should use $T$ instead of $P$ for period because
$P$ is too heavily overloaded by the probabilities\ldots}

We have some parametric function \rate\ describing the true rate of exoplanets
with periods \period\ and radii \radius
\begin{eqnarray}\eqlabel{true-rate}
\rate (\period,\,\radius;\,\ratepars)
&=& \frac{\dd N}{\dd\ln\period\dd\ln\radius}\quad.
\end{eqnarray}
Here we have not made any assumptions about the functional form of \rate, we
are only stating that it can be modeled by some---possibly very large---set of
parameters \ratepars.

In this \paper, we will assume that the completeness function of the survey is
known and it can be represented as a probability at every period and radius
$\completeness (\period,\,\radius)$.
Similarly, we can consider the marginalized transit probability $\transitprob
(\period)$ as only a function of period.
These selection effects can be incorporated into the rate from \eq{true-rate}
to get the \emph{observed rate}
\begin{eqnarray}\eqlabel{obs-rate}
\obs{\rate} (\period,\,\radius;\,\ratepars)
&=& \rate (\period,\,\radius;\,\ratepars)\,\completeness(\period,\,\radius)\,
\transitprob(\period)\quad.
\end{eqnarray}

Now we will make the further assumption that every candidate in our sample of
detected exoplanets is an independent Poisson sample from the population
described by our rate function.
Under this assumption, the likelihood of the catalog conditioned on a set of
parameters \ratepars\ is
\begin{eqnarray}
\ln p(\period,\,\radius\,|\,\ratepars) &\propto&
\sum_{k=1}^K \ln\obs{\rate} (\period_k,\,\radius_k;\,\ratepars)
- \int \obs{\rate} (\period,\,\radius;\,\ratepars) \dd\period\dd\radius
\quad.
\end{eqnarray}

\hoggtodo{How is this related to the version of the likelihood that doesn't
include the normalization. What is the best way to introduce both forms?}

\section{Including observational uncertainties}

In the previous section, we implicitly assumed that the measurements of
\period\ and \radius\ given by the catalog were perfect, without any
observational uncertainties.
This isn't a terrible approximation for the period measurements but not for
the radii.
Luckily, it is not hard to include these effects in our probabilistic
framework.

In general, catalogs produce posterior probability distributions for
quantities conditioned on data.
In particular, the catalog that we're considering here provides estimates of
the posterior distribution of \period\ and \radius\ conditioned on the
observables $\obs{\period}$ and $\obs{\radius}$ under some \emph{interim
priors} parameterized by \interim
\begin{eqnarray}
p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim) \quad.
\end{eqnarray}
This estimate can be propagated as \emph{posterior samples}
\begin{eqnarray}
\{\period^{(n)},\,\radius^{(n)}\} &\sim&
p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim) \quad.
\end{eqnarray}

Now the quantity that we need to compute is again the (marginalized)
likelihood of the data $\obs{\period}$ and $\obs{\radius}$ conditioned on a
set of rate function parameters \ratepars
\begin{eqnarray}
p(\obs{\period},\,\obs{\radius}\,|\,\ratepars) &=&
\int p(\period,\,\radius,\,\obs{\period},\,\obs{\radius}\,|\,\ratepars)
\dd\period\dd\radius \\
&=&
\int p(\period,\,\radius\,|\,\ratepars) \,
p(\obs{\period},\,\obs{\radius}\,|\,\period,\,\radius)
\dd\period\dd\radius \\
&=&
\int p(\period,\,\radius\,|\,\ratepars) \,
p(\obs{\period},\,\obs{\radius}\,|\,\period,\,\radius)\,
\frac{p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim)}
     {p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim)}
\dd\period\dd\radius \\
&=&
p(\obs{\period},\,\obs{\radius}\,|\,\interim)\,
\int
\frac{p(\period,\,\radius\,|\,\ratepars) \,
      p(\obs{\period},\,\obs{\radius}\,|\,\period,\,\radius)}
     {p(\period,\,\radius\,|\,\interim) \,
      p(\obs{\period},\,\obs{\radius}\,|\,\period,\,\radius)}\,
p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim)
\dd\period\dd\radius \\
\frac{p(\obs{\period},\,\obs{\radius}\,|\,\ratepars)}
     {p(\obs{\period},\,\obs{\radius}\,|\,\interim)}
&=&
\int
\frac{p(\period,\,\radius\,|\,\ratepars)}
     {p(\period,\,\radius\,|\,\interim)} \,
p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim)
\dd\period\dd\radius \\
&\approx&
\frac{1}{N}
\sum_{n=1}^N
\frac{p(\period^{(n)},\,\radius^{(n)}\,|\,\ratepars)}
     {p(\period^{(n)},\,\radius^{(n)}\,|\,\interim)}
\end{eqnarray}

\acknowledgments
It is a pleasure to thank
    Erik Petigura (Berkeley),
    \ldots
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).

@@

\bibitem[Petigura \etal(2013)]{petigura}
Petigura, E.~A., Howard, A.~W., \& Marcy, G.~W.\ 2013,
Proceedings of the National Academy of Science, 110, 19273
